---
title: "Initial QC & Data Cleaning — Dom Proteomics (SORL1 xenograft microglia)"
---

## Libraries

```{r setup, include=TRUE, message=FALSE}
library(data.table)
library(dplyr)
library(ggplot2)
library(stringr)
library(tidyr)
library(pheatmap)
library(ggrepel)
```

## Set directories

```{r}
base_dir    <- "/nemo/lab/destrooperb/home/shared/zanettc/dom_proteomics"
run_num     <- "run1"
input_dir   <- file.path(base_dir, "data", "inputs",
                         "2026_04_23_SORL1_mg_total_proteomics")
input_file  <- file.path(input_dir, "20260423_094931_S5-A10_Report_human.tsv")

results_dir <- file.path(base_dir, "results", run_num)
objects_dir <- file.path(base_dir, "data", "processed", run_num)

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(objects_dir, recursive = TRUE, showWarnings = FALSE)
```

## Define sample metadata

8 samples from a single S5 mass-spec batch (EXP1.1 cohort). Human iPSC-derived microglia xenografted into App NL-G-F mice; SORL1 genotype WT or KO. HET samples (ds31/32/41/42, EXP1.2) are not present in this file.

Note: BDAG173.1f was labelled ds52 in the study metadata but recorded as ds51 in Spectronaut. It is distinguished here by its plate position (S5-A10).

```{r}
meta <- data.table(
  FileName  = c(
    "ds12_S5-A1_1_6411",   "ds11_S5-A2_1_6412",
    "ds22_S5-A3_1_6413",   "ds21_S5-A4_1_6414",
    "ds51_S5-A9_1_6421",   "ds51_S5-A10_1_6422",
    "ds61_S5-A11_1_6423",  "ds62_S5-A12_1_6424"
  ),
  ShortName = c(
    "BDAG171.1c", "BDAG171.1d",
    "BDAG171.1a", "BDAG171.1b",
    "BDAG173.1e", "BDAG173.1f",
    "BDAG173.1b", "BDAG173.1a"
  ),
  SampleNum = 1:8,
  Genotype  = c("WT", "WT", "WT", "WT", "KO", "KO", "KO", "KO"),
  Sex       = c("Male", "Male", "Male", "Male", "Female", "Female", "Male", "Male"),
  CellCount = c(154300, 143900, 463400, 69340, 141800, 209800, 95890, 237400)
)
meta
```

## Read Spectronaut report

```{r}
cat("Reading Spectronaut report (protein-level columns only)...\n")
spec_raw <- fread(
  input_file,
  select = c("R.FileName", "PG.ProteinGroups", "PG.ProteinAccessions",
             "PG.Qvalue", "PG.Quantity")
)

cat("Fragment rows loaded:", nrow(spec_raw), "\n")
cat("Unique samples in file:", uniqueN(spec_raw$R.FileName), "\n")

spec_dom <- spec_raw[R.FileName %in% meta$FileName]
cat("Fragment rows for Dom's samples:", nrow(spec_dom), "\n")
rm(spec_raw); gc()
```

## Extract unique protein-level quantities

PG.Quantity repeats for every fragment row of a protein within a sample — deduplicate to one row per protein–sample pair.

```{r}
pg_long <- unique(spec_dom[, .(R.FileName, PG.ProteinGroups, PG.ProteinAccessions,
                                PG.Qvalue, PG.Quantity)])

cat("Unique protein-sample rows:", nrow(pg_long), "\n")
cat("Unique proteins:           ", uniqueN(pg_long$PG.ProteinGroups), "\n")
```

## Pivot to wide format (proteins × samples)

Because BDAG173.1e and BDAG173.1f share the R.FileName prefix "ds51", we join via meta (by plate position / FileName) before pivoting so each column carries a unique ShortName.

```{r}
pg_long_named <- merge(pg_long, meta[, .(FileName, ShortName)],
                       by.x = "R.FileName", by.y = "FileName")

quant_wide <- dcast(
  pg_long_named[, .(ShortName, PG.ProteinGroups, PG.Quantity)],
  PG.ProteinGroups ~ ShortName,
  value.var = "PG.Quantity"
)

sample_cols <- meta$ShortName

missing_samples <- setdiff(sample_cols, colnames(quant_wide))
if (length(missing_samples) > 0) {
  warning("Missing from data: ", paste(missing_samples, collapse = ", "))
} else {
  cat("All", length(sample_cols), "samples found in wide matrix.\n")
}

quant_data   <- quant_wide[, ..sample_cols]
quant_matrix <- as.matrix(quant_data)
quant_matrix[is.nan(quant_matrix)] <- NA
colnames(quant_matrix) <- meta$ShortName

cat("Proteins in matrix:", nrow(quant_matrix), "\n")
cat("Samples in matrix: ", ncol(quant_matrix), "\n")
```

## Protein ID counts and % missing per sample

With only 8 samples, statistical outlier rules (e.g. 3-MAD) are unreliable — use protein ID count and % missing as complementary metrics. Cell count is shown alongside, as very low input material (BDAG171.1b, 69 K cells) may drive missingness independently of technical QC failure.

```{r}
total_pool     <- nrow(quant_matrix)
protein_counts <- colSums(!is.na(quant_matrix) & quant_matrix > 0)
na_pct         <- colMeans(is.na(quant_matrix) | quant_matrix == 0) * 100

qc_counts <- data.table(
  ShortName     = names(protein_counts),
  Protein_Count = protein_counts,
  NA_Percent    = na_pct[names(protein_counts)]
)
qc_data <- merge(qc_counts, meta, by = "ShortName", all.x = TRUE)

cat("Total protein pool (union across all samples):", total_pool, "\n\n")
summary(qc_data$Protein_Count)
qc_data[order(NA_Percent), .(ShortName, Genotype, Sex, CellCount,
                              Protein_Count, NA_Percent = round(NA_Percent, 1))]
```

## Flag samples for review

```{r}
miss_threshold <- 60

flagged_runs <- qc_data[NA_Percent > miss_threshold, ShortName]

if (length(flagged_runs) > 0) {
  cat("Samples above", miss_threshold, "% missing (flagged for review):\n")
  print(flagged_runs)
} else {
  cat("No samples exceed the", miss_threshold, "% missing threshold.\n")
}

# Update failed_runs after reviewing the QC plots below.
failed_runs <- character(0)
```

## Peak quality: area / height ratio proxy

`EG.FWHM` is not present in this export — as a proxy we use NormalizedPeakArea / NormalizedPeakHeight at the fragment level (proportional to peak width). A systematic upward shift in one sample flags poor chromatographic performance.

```{r}
cat("Reading fragment-level columns for peak quality...\n")
frag_raw <- fread(
  input_file,
  select = c("R.FileName", "EG.PrecursorId",
             "F.NormalizedPeakArea", "F.NormalizedPeakHeight",
             "F.ExcludedFromQuantification")
)

frag_dom <- frag_raw[R.FileName %in% meta$FileName]
rm(frag_raw); gc()

frag_dom <- merge(frag_dom, meta[, .(FileName, ShortName, Genotype, Sex, SampleNum)],
                  by.x = "R.FileName", by.y = "FileName")

frag_dom <- frag_dom[
  F.ExcludedFromQuantification == FALSE &
    !is.na(F.NormalizedPeakArea) &
    F.NormalizedPeakHeight > 0
]

frag_dom[, Peak_Width_Proxy := F.NormalizedPeakArea / F.NormalizedPeakHeight]

cat("Quantified fragments retained:", nrow(frag_dom), "\n")
cat("Samples present:              ", uniqueN(frag_dom$ShortName), "\n")
```

```{r, fig.width=10, fig.height=5}
genotype_colours <- c("WT" = "#1b9e77", "KO" = "#d95f02")
sex_shapes       <- c("Male" = 16, "Female" = 17)

p_peak <- ggplot(frag_dom,
                 aes(x = reorder(ShortName, SampleNum[match(ShortName, meta$ShortName)]),
                     y = Peak_Width_Proxy, fill = Genotype)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  coord_cartesian(ylim = quantile(frag_dom$Peak_Width_Proxy, c(0.01, 0.99), na.rm = TRUE)) +
  scale_fill_manual(values = genotype_colours) +
  theme_minimal() +
  labs(title = "Peak Width Proxy per Sample (NormalisedArea / NormalisedHeight)",
       subtitle = "Higher = broader peaks. Proper FWHM requires re-export with EG.FWHM column.",
       x = "Sample", y = "Area / Height ratio") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p_peak)
```

```{r, fig.width=8, fig.height=5}
peak_summary <- frag_dom[, .(
  Median_Width_Proxy = median(Peak_Width_Proxy, na.rm = TRUE),
  P25 = quantile(Peak_Width_Proxy, 0.25, na.rm = TRUE),
  P75 = quantile(Peak_Width_Proxy, 0.75, na.rm = TRUE)
), by = .(ShortName, Genotype, Sex)]

peak_summary <- merge(peak_summary, qc_data[, .(ShortName, SampleNum)], by = "ShortName")

p_peak_dot <- ggplot(peak_summary,
                     aes(x = reorder(ShortName, SampleNum),
                         y = Median_Width_Proxy, color = Genotype, shape = Sex)) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = P25, ymax = P75), width = 0.3) +
  scale_color_manual(values = genotype_colours) +
  scale_shape_manual(values = sex_shapes) +
  theme_minimal() +
  labs(title = "Median Peak Width Proxy ± IQR",
       subtitle = "Samples far above the group median have broader / lower-quality peaks",
       x = "Sample", y = "Median Area / Height ratio") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p_peak_dot)
```

## Generate QC report

```{r, fig.width=10, fig.height=7}
# ── Plot 1: Protein Count per Sample ───────────────────────────────────────
p1 <- ggplot(qc_data,
             aes(x = reorder(ShortName, Protein_Count),
                 y = Protein_Count, fill = Genotype)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = genotype_colours) +
  coord_flip() +
  theme_minimal() +
  labs(title = "Total Proteins Identified per Sample",
       x = "Sample", y = "Protein Count")

# ── Plot 2: % Missing per Sample ────────────────────────────────────────────
p2 <- ggplot(qc_data,
             aes(x = reorder(ShortName, NA_Percent),
                 y = NA_Percent, fill = Genotype)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = miss_threshold, color = "red", linetype = "dashed", linewidth = 1) +
  annotate("text", x = 1.5, y = miss_threshold + 1,
           label = paste0(miss_threshold, "% threshold"), color = "red", hjust = 0) +
  scale_fill_manual(values = genotype_colours) +
  coord_flip() +
  theme_minimal() +
  labs(title = "% Missing Values per Sample",
       subtitle = "% of the total protein pool absent in each sample",
       x = "Sample", y = "% Missing")

# ── Plot 3: Protein Yield by Genotype with cell count ───────────────────────
p3 <- ggplot(qc_data, aes(x = Genotype, y = Protein_Count, fill = Genotype)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_point(aes(shape = Sex, size = CellCount / 1e3), alpha = 0.8,
             position = position_jitter(width = 0.15)) +
  geom_text_repel(aes(label = ShortName), size = 3) +
  scale_fill_manual(values = genotype_colours) +
  scale_shape_manual(values = sex_shapes) +
  scale_x_discrete(limits = c("WT", "KO")) +
  theme_minimal() +
  labs(title = "Protein Yield by Genotype",
       subtitle = "Point size = cell count (thousands); shape = sex",
       x = "Genotype", y = "Protein Count", size = "Cell count (K)")

# ── PCA ─────────────────────────────────────────────────────────────────────
cat("Preparing PCA...\n")

pca_mat_log   <- log2(quant_matrix + 1)
pca_mat_log_t <- t(pca_mat_log)

noise_floor <- min(pca_mat_log_t[pca_mat_log_t > 0], na.rm = TRUE) / 2
pca_mat_log_t[is.na(pca_mat_log_t) | pca_mat_log_t == 0] <- noise_floor

valid_cols    <- apply(pca_mat_log_t, 2, var) > 0
pca_mat_log_t <- pca_mat_log_t[, valid_cols]
cat("Proteins retained for PCA (non-zero variance):", sum(valid_cols), "\n")

pca_res  <- prcomp(pca_mat_log_t, center = TRUE, scale. = TRUE)
var_expl <- round(100 * pca_res$sdev^2 / sum(pca_res$sdev^2), 1)

pca_df <- as.data.frame(pca_res$x)
pca_df$ShortName <- rownames(pca_df)
pca_df <- merge(pca_df, meta, by = "ShortName", all.x = TRUE)
pca_df <- merge(pca_df, qc_data[, .(ShortName, NA_Percent)], by = "ShortName")

# ── Plot 4: PCA by Genotype ─────────────────────────────────────────────────
p4 <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Genotype, shape = Sex)) +
  geom_point(size = 4, alpha = 0.85) +
  geom_text_repel(aes(label = ShortName), size = 3, max.overlaps = 20) +
  scale_color_manual(values = genotype_colours) +
  scale_shape_manual(values = sex_shapes) +
  theme_minimal() +
  labs(title = "PCA of Log2 Protein Intensities",
       subtitle = "Color = SORL1 genotype; shape = sex",
       x = paste0("PC1 (", var_expl[1], "%)"),
       y = paste0("PC2 (", var_expl[2], "%)"))

# ── Plot 5: PCA coloured by % missing ───────────────────────────────────────
p5 <- ggplot(pca_df, aes(x = PC1, y = PC2, color = NA_Percent)) +
  geom_point(size = 4, alpha = 0.85) +
  geom_text_repel(aes(label = ShortName), size = 3, max.overlaps = 20) +
  scale_color_viridis_c(name = "% Missing", option = "plasma") +
  theme_minimal() +
  labs(title = "PCA coloured by % Missing",
       subtitle = "If a weak sample drives a PC, it should be removed before MSstats",
       x = paste0("PC1 (", var_expl[1], "%)"),
       y = paste0("PC2 (", var_expl[2], "%)"))

# ── Plot 6: PC3 vs PC4 ──────────────────────────────────────────────────────
p6 <- ggplot(pca_df, aes(x = PC3, y = PC4, color = Genotype, shape = Sex)) +
  geom_point(size = 4, alpha = 0.85) +
  geom_text_repel(aes(label = ShortName), size = 3, max.overlaps = 20) +
  scale_color_manual(values = genotype_colours) +
  scale_shape_manual(values = sex_shapes) +
  theme_minimal() +
  labs(title = "PCA — PC3 vs PC4",
       x = paste0("PC3 (", var_expl[3], "%)"),
       y = paste0("PC4 (", var_expl[4], "%)"))

# ── Save PDF ────────────────────────────────────────────────────────────────
pdf_path <- file.path(results_dir, "Initial_QC_Report.pdf")
cat("Saving PDF:", pdf_path, "\n")
pdf(pdf_path, width = 10, height = 7)
print(p1); print(p2); print(p3); print(p4); print(p5); print(p6)
print(p_peak); print(p_peak_dot)
dev.off()

print(p1); print(p2); print(p3); print(p4); print(p5); print(p6)
```

## Abundance distributions

Log2 intensity boxplots per sample, ordered by sample number. Flagged samples shown with a red outline.

```{r, fig.width=10, fig.height=6}
quant_log_plot <- log2(quant_matrix + 1)
quant_log_plot[quant_log_plot == 0] <- NA

abund_long <- as.data.frame(quant_log_plot) %>%
  mutate(Protein = row_number()) %>%
  pivot_longer(-Protein, names_to = "ShortName", values_to = "Log2_Intensity") %>%
  drop_na() %>%
  left_join(as.data.frame(qc_data), by = "ShortName") %>%
  mutate(Status = ifelse(ShortName %in% failed_runs, "Flagged", "Pass"))

p_abund <- ggplot(abund_long, aes(x = reorder(ShortName, SampleNum),
                                   y = Log2_Intensity, fill = Genotype)) +
  geom_boxplot(aes(color = Status), outlier.shape = NA, lwd = 0.4) +
  scale_fill_manual(values = genotype_colours) +
  scale_color_manual(values = c("Pass" = "black", "Flagged" = "red")) +
  theme_minimal() +
  labs(title = "Log2 Abundance per Sample",
       subtitle = "Red outlines = samples exceeding % missing threshold",
       x = "Sample", y = "Log2 Intensity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p_abund)
```

## Correlation heatmap

Samples grouped by genotype then sex so biological structure is visible.

```{r, fig.width=8, fig.height=7}
quant_mat_cor <- log2(quant_matrix + 1)
quant_mat_cor[quant_mat_cor == 0] <- NA

cor_mat <- cor(quant_mat_cor, use = "pairwise.complete.obs")

meta_sorted    <- meta[order(Genotype, Sex, SampleNum)] %>% as.data.frame()
sorted_names   <- meta_sorted$ShortName
cor_mat_sorted <- cor_mat[sorted_names, sorted_names]

anno_df <- meta_sorted %>%
  select(ShortName, Genotype, Sex) %>%
  as.data.frame()
rownames(anno_df) <- anno_df$ShortName
anno_df$ShortName <- NULL

pheatmap(cor_mat_sorted,
         main              = "Pearson Correlation: Grouped by Genotype × Sex",
         annotation_col    = anno_df,
         annotation_colors = list(
           Genotype = genotype_colours,
           Sex      = c("Male" = "#4393c3", "Female" = "#d6604d")
         ),
         cluster_rows      = FALSE,
         cluster_cols      = FALSE,
         show_rownames     = TRUE,
         show_colnames     = TRUE,
         color             = colorRampPalette(c("navy", "white", "firebrick3"))(100),
         gaps_col          = which(diff(as.numeric(as.factor(meta_sorted$Genotype))) != 0),
         gaps_row          = which(diff(as.numeric(as.factor(meta_sorted$Genotype))) != 0))
```

## Average-correlation diagnostic

Identifies the sample with the lowest mean pairwise correlation.

```{r}
qc_diagnostics <- data.table(
  ShortName       = colnames(quant_mat_cor),
  Protein_Count   = protein_counts[colnames(quant_mat_cor)],
  NA_Percent      = na_pct[colnames(quant_mat_cor)],
  Avg_Correlation = rowMeans(cor_mat, na.rm = TRUE)
)

outlier_id <- qc_diagnostics[which.min(Avg_Correlation), ShortName]
cat("--- Lowest-correlation sample ---\n")
print(qc_diagnostics[ShortName == outlier_id])
cat("\nAll samples ranked by correlation:\n")
print(qc_diagnostics[order(Avg_Correlation)])

p_diag <- ggplot(qc_diagnostics, aes(x = Protein_Count, y = Avg_Correlation)) +
  geom_point(aes(color = NA_Percent), size = 4) +
  geom_text_repel(aes(label = ShortName), size = 3, max.overlaps = 20) +
  scale_color_viridis_c(name = "% Missing", option = "plasma") +
  theme_minimal() +
  labs(title = "Signal vs. Correlation Diagnostic",
       subtitle = "Color = % missing; samples bottom-left are weakest",
       x = "Total Protein IDs", y = "Average Pearson Correlation (r)")

print(p_diag)
```

## Contaminant check: top-10 protein signal contribution

Values \>50% suggest a very low-complexity sample or dominant contaminants (e.g. keratins, actin).

```{r}
calc_top_n_pct <- function(column, n = 10) {
  sorted_vals <- sort(column, decreasing = TRUE, na.last = TRUE)
  (sum(sorted_vals[1:n], na.rm = TRUE) / sum(column, na.rm = TRUE)) * 100
}

top10_metrics <- apply(quant_matrix, 2, calc_top_n_pct, n = 10)
qc_data[, Top10_Percent   := top10_metrics[ShortName]]
qc_data[, Total_Intensity := colSums(quant_matrix, na.rm = TRUE)[ShortName]]

contaminant_suspects <- qc_data[Top10_Percent > 50,
                                 .(ShortName, Genotype, Sex, Protein_Count, Top10_Percent)]
cat("Samples where top 10 proteins account for >50% of signal:\n")
print(contaminant_suspects)

p_top10 <- ggplot(qc_data, aes(x = Protein_Count, y = Top10_Percent)) +
  geom_point(aes(color = Genotype, shape = Sex, size = Total_Intensity), alpha = 0.7) +
  geom_hline(yintercept = 50, linetype = "dashed", color = "red") +
  geom_text_repel(
    aes(label = ifelse(Top10_Percent > 50, ShortName, "")),
    size = 3, max.overlaps = 10
  ) +
  scale_color_manual(values = genotype_colours) +
  scale_shape_manual(values = sex_shapes) +
  theme_minimal() +
  labs(title = "Top-10 Protein Signal Contribution",
       subtitle = ">50% suggests contamination or extremely low sample complexity",
       x = "Total Proteins Identified",
       y = "% Total Intensity from Top 10 Proteins",
       size = "Total Signal (raw)")

print(p_top10)
```

## Identify top contributing proteins in suspect samples

```{r}
if (nrow(contaminant_suspects) > 0) {
  suspect_short <- contaminant_suspects$ShortName
  top_ids <- apply(quant_matrix[, suspect_short, drop = FALSE], 2, function(x) {
    quant_wide$PG.ProteinGroups[which.max(x)]
  })
  cat("Top protein per suspect sample:\n")
  print(top_ids)
} else {
  cat("No suspect samples to investigate.\n")
}
```

## Conclusions

```{r}
cat("=== QC SUMMARY ===\n\n")
cat("Total samples:                       ", nrow(meta), "\n")
cat("Samples dropped (manual review):     ", length(failed_runs), "\n")
if (length(failed_runs) > 0) {
  cat("  -> Dropped:", paste(failed_runs, collapse = ", "), "\n")
}

final_names <- setdiff(meta$ShortName, failed_runs)
cat("Samples retained:                    ", length(final_names), "\n\n")

qc_data[, .(ShortName, Genotype, Sex, CellCount, Protein_Count,
            NA_Percent      = round(NA_Percent, 1),
            Avg_Correlation = round(rowMeans(cor_mat, na.rm = TRUE), 3),
            Top10_Percent   = round(Top10_Percent, 1))][order(Genotype, Sex, ShortName)]
```

## Save objects

```{r}
final_filenames   <- meta[ShortName %in% final_names, FileName]
final_short_names <- meta[FileName %in% final_filenames, ShortName]

clean_spec_dom <- spec_dom[R.FileName %in% final_filenames]

clean_quant_wide <- quant_wide[, c("PG.ProteinGroups", final_short_names), with = FALSE]

clean_meta <- qc_data[ShortName %in% final_names]

if (length(failed_runs) > 0) {
  dropped_log <- qc_data[ShortName %in% failed_runs,
                          .(ShortName, FileName, Genotype, Sex, Protein_Count, NA_Percent)]
  dropped_log[, Reason := paste0("Above ", miss_threshold, "% missing threshold")]
  fwrite(dropped_log, file.path(objects_dir, "dropped_samples_log.csv"))
}

fwrite(clean_spec_dom,   file.path(objects_dir, "clean_spec_dom.csv"))
fwrite(clean_meta,       file.path(objects_dir, "clean_metadata.csv"))
fwrite(clean_quant_wide, file.path(objects_dir, "clean_quant_wide.csv"))

cat("Objects saved to:", objects_dir, "\n")
```
