---
title: "2. MSstats — Differential Abundance Analysis"
---

## Libraries

```{r setup, include=TRUE, message=FALSE}
library(data.table)
library(MSstats)
library(qs2)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(tidyverse)
library(readr)
library(AnnotationDbi)
library(org.Hs.eg.db)
```

## Set directories

```{r}
base_dir    <- "/nemo/lab/destrooperb/home/shared/zanettc/dom_proteomics"
run_num     <- "run2"

results_dir <- file.path(base_dir, "results", run_num)
objects_dir <- file.path(base_dir, "data", "processed", run_num)
input_dir   <- file.path(base_dir, "data", "inputs",
                         "2026_04_23_SORL1_mg_total_proteomics")

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(objects_dir, recursive = TRUE, showWarnings = FALSE)
```

## Build annotation from study design

All 8 samples passed QC. R.Condition in the raw file is "Not Defined" — we assign
WT/KO here from the known sample table.

Note: WT samples are all Male (BDAG171 cohort); KO samples are 2 Female + 2 Male
(BDAG173 cohort). Sex is not modelled as a covariate in the MSstats linear model —
the KO vs WT contrast is therefore partially confounded with sex.

```{r}
msstats_annotation <- data.table(
  Run = c(
    "ds12_S5-A1_1_6411", "ds11_S5-A2_1_6412",
    "ds22_S5-A3_1_6413", "ds21_S5-A4_1_6414",
    "ds51_S5-A9_1_6421", "ds51_S5-A10_1_6422",
    "ds61_S5-A11_1_6423", "ds62_S5-A12_1_6424"
  ),
  Condition    = c(rep("WT", 4), rep("KO", 4)),
  BioReplicate = 1:8
)

msstats_annotation

fwrite(msstats_annotation, file.path(objects_dir, "msstats_annotation.csv"))
```

## Load raw Spectronaut report

Select only the fragment-level columns required by MSstats to keep memory manageable.

```{r}
msstats_cols <- c(
  "R.Condition", "R.FileName", "R.Replicate",
  "PG.ProteinGroups", "PG.ProteinAccessions",
  "PEP.StrippedSequence",
  "EG.ModifiedSequence", "EG.PrecursorId", "EG.Qvalue",
  "FG.Charge", "FG.Id",
  "F.FrgIon", "F.Charge", "F.FrgLossType", "F.FrgNum", "F.FrgType",
  "F.NormalizedPeakArea", "F.PeakArea", "F.ExcludedFromQuantification"
)

cat("Reading Spectronaut report...\n")
raw_spectronaut <- fread(
  file.path(input_dir, "20260423_094931_S5-A10_Report_human.tsv"),
  select = msstats_cols
)
cat("Rows loaded:", nrow(raw_spectronaut), "\n")
```

## Filter to human proteins

The Spectronaut search used a concatenated human + mouse FASTA (see notebook 04).
Retain only rows whose primary protein group accession maps to `org.Hs.eg.db` so
that mouse host-derived proteins are excluded before MSstats sees the data.

```{r}
prim_acc  <- str_extract(raw_spectronaut$PG.ProteinGroups, "^[^;]+")
human_map <- mapIds(org.Hs.eg.db, unique(prim_acc),
                   "SYMBOL", "UNIPROT", multiVals = "first")
human_acc <- names(human_map)[!is.na(human_map)]

n_pg_before <- uniqueN(raw_spectronaut$PG.ProteinGroups)
raw_spectronaut <- raw_spectronaut[prim_acc %in% human_acc]
n_pg_after  <- uniqueN(raw_spectronaut$PG.ProteinGroups)

cat("Protein groups before human filter:", n_pg_before, "\n")
cat("Protein groups after  human filter:", n_pg_after,  "\n")
cat("Mouse protein groups removed:      ", n_pg_before - n_pg_after, "\n")

qs_save(raw_spectronaut, file.path(objects_dir, "msstats_report.qs2"))
```

## Reload from .qs2 (skip read on re-runs)

```{r eval=FALSE}
raw_spectronaut <- qs_read(file.path(objects_dir, "msstats_report.qs2"))
```

## Overwrite R.Condition and R.Replicate with study design values

```{r}
raw_spectronaut[msstats_annotation,
                on = c("R.FileName" = "Run"),
                `:=`(R.Condition = i.Condition,
                     R.Replicate = i.BioReplicate)]

cat("Condition distribution after fix:\n")
print(raw_spectronaut[, .N, by = .(R.Condition, R.FileName)][order(R.Condition, R.FileName)])
```

## Convert to MSstats input format

Feature-level filtering: Qvalue < 0.01, remove features with too few measurements.

```{r}
msstats_input <- suppressWarnings(SpectronauttoMSstatsFormat(
  raw_spectronaut,
  annotation                 = msstats_annotation,
  filter_with_Qvalue         = TRUE,
  qvalue_cutoff              = 0.01,
  fewMeasurements            = "remove",
  removeProtein_with1Feature = FALSE
))

cat("MSstats input rows:", nrow(msstats_input), "\n")
cat("Unique proteins:   ", uniqueN(msstats_input$ProteinName), "\n")

qs_save(msstats_input, file.path(objects_dir, "msstats_input.qs2"))
```

## Feature completeness filter

Keep features observed in ≥50% of runs within at least one condition
(≥ 2 of 4 samples in WT or ≥ 2 of 4 samples in KO).

```{r eval=FALSE}
# Manual resume — skip rebuild on re-run
msstats_input <- qs_read(file.path(objects_dir, "msstats_input.qs2"))
```

```{r}
msstats_input_filt <- as.data.frame(msstats_input) %>%
  group_by(ProteinName, PeptideSequence, PrecursorCharge,
           FragmentIon, ProductCharge, Condition) %>%
  summarise(
    n_obs   = sum(!is.na(Intensity) & Intensity > 0),
    n_total = n(),
    prop    = n_obs / n_total,
    .groups = "drop"
  ) %>%
  group_by(ProteinName, PeptideSequence, PrecursorCharge,
           FragmentIon, ProductCharge) %>%
  filter(any(prop >= 0.5)) %>%
  ungroup() %>%
  inner_join(
    as.data.frame(msstats_input),
    by = c("ProteinName", "PeptideSequence", "PrecursorCharge",
           "FragmentIon", "ProductCharge", "Condition")
  )

message("Rows before completeness filter: ", nrow(msstats_input))
message("Rows after  completeness filter: ", nrow(msstats_input_filt))
message("Proteins retained: ", uniqueN(msstats_input_filt$ProteinName))
```

## Data summarisation and normalisation

Summarises fragment intensities to protein level with TMP, median normalisation,
and model-based imputation for missing values.

```{r}
processed_data <- suppressWarnings(dataProcess(
  msstats_input_filt,
  normalization                        = "equalizeMedians",
  summaryMethod                        = "TMP",
  censoredInt                          = "NA",
  MBimpute                             = TRUE,
  featureSubset                        = "highQuality",
  remove_uninformative_feature_outlier = TRUE
))

qs_save(processed_data, file.path(objects_dir, "processed_msstats_data.qs2"))
```

## Reload processed data (if re-running from this point)

```{r eval=FALSE}
processed_data <- qs_read(file.path(objects_dir, "processed_msstats_data.qs2"))
```

## Verify condition levels

MSstats sorts conditions alphabetically — confirm the order before building the
contrast matrix. Expected: "KO" then "WT".

```{r}
condition_levels <- levels(processed_data$ProteinLevelData$GROUP)
cat("Condition levels (alphabetical):\n")
print(condition_levels)
```

## Build contrast matrix

Single comparison: KO vs WT. Positive log2FC = higher abundance in KO.

```{r}
make_contrast <- function(numerator, denominator, levels) {
  v <- setNames(rep(0L, length(levels)), levels)
  v[numerator]   <-  1L
  v[denominator] <- -1L
  v
}

comparison_matrix <- rbind(
  make_contrast("KO", "WT", condition_levels)
)
rownames(comparison_matrix) <- "KO_vs_WT"

print(comparison_matrix)
```

## Statistical modelling

Linear mixed-effects model across all proteins.

```{r}
test_results <- suppressWarnings(
  groupComparison(contrast.matrix = comparison_matrix, data = processed_data)
)

qs_save(test_results, file.path(objects_dir, "comparison_results.qs2"))

res_df <- as.data.table(test_results$ComparisonResult)
cat("Total protein-comparison rows:", nrow(res_df), "\n")
cat("Significant (adj.pvalue < 0.05):",
    nrow(res_df[adj.pvalue < 0.05, ]), "\n")

fwrite(res_df, file.path(results_dir, "Full_GroupComparison_Results.csv"))
```

## Summary of results

```{r}
res_df %>%
  filter(!is.na(adj.pvalue)) %>%
  group_by(Label) %>%
  summarise(
    n_tested = n(),
    n_sig    = sum(adj.pvalue < 0.05, na.rm = TRUE),
    n_up     = sum(adj.pvalue < 0.05 & log2FC > 0, na.rm = TRUE),
    n_down   = sum(adj.pvalue < 0.05 & log2FC < 0, na.rm = TRUE),
    .groups  = "drop"
  )
```

## Reload comparison results (if re-running from this point)

```{r eval=FALSE}
test_results <- qs_read(file.path(objects_dir, "comparison_results.qs2"))
res_df       <- as.data.table(test_results$ComparisonResult)
```
