---
title: "4. FASTA Audit — Mislabelled Search Database"
---

## TL;DR

The Spectronaut report delivered as `20260423_094931_S5-A10_Report_human.tsv` was searched against a FASTA file named `20K_HUMAN_1GENE_202210.fasta` — implying a human-only proteome database. **The output contains \~50 % mouse-origin proteins**, which is incompatible with a strictly human FASTA: UniProt accessions are globally unique across organisms, so a clean mouse accession in the output proves the FASTA contains mouse sequences.

The likely facts: 1. The FASTA file is a concatenated **human + mouse** database 2. The filename suggests human-only 3. Without species filtering, every downstream analysis is contaminated by host (mouse) proteins — this doubles the multiple-testing burden, distorts median normalisation, and confounds the WT vs KO contrast with sample-to-sample host contamination.

## Setup

```{r setup, message=FALSE}
library(data.table)
library(dplyr)
library(stringr)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(org.Mm.eg.db)

base_dir    <- "/nemo/lab/destrooperb/home/shared/zanettc/dom_proteomics"
input_dir   <- file.path(base_dir, "data", "inputs",
                         "2026_04_23_SORL1_mg_total_proteomics")
results_dir <- file.path(base_dir, "results", "run1")
```

## Evidence 1 — Filename and stated database

The delivered report is named **`20260423_094931_S5-A10_Report_human.tsv`** with sibling files `*.params`, `*.setup.txt`, `*.log.txt`. The setup file declares the search database used.

```{r, warning=FALSE, message=FALSE}
setup_path <- file.path(input_dir, "20260423_094931_S5-A10_Report_human.setup.txt")

fasta_lines <- system2(
  "grep", c("-A1", "Original File", setup_path),
  stdout = TRUE
)
unique(fasta_lines[grepl("Original File", fasta_lines)])
```

The setup file states a single search database: `20K_HUMAN_1GENE_202210.fasta`. By name alone, this implies a Homo sapiens proteome with one canonical isoform per gene (\~20 000 entries).

## Evidence 2 — UniProt accession identity is species-specific

UniProt accessions are **globally unique across organisms** — the same accession never refers to two proteins in two species. Therefore, if the output of a search against a strictly human FASTA contains an accession that, in the UniProt knowledge base, is annotated to a non-human organism, the FASTA cannot have been strictly human.

We test this against the full set of unique protein groups in the MSstats output by mapping the primary accession of each group to (a) `org.Hs.eg.db` and (b) `org.Mm.eg.db`.

```{r, warning=FALSE, message=FALSE}
res <- fread(file.path(results_dir, "Full_GroupComparison_Results.csv"))
prim <- str_extract(res$Protein, "^[^;]+")  # first accession in semicolon-separated group

hs <- !is.na(mapIds(org.Hs.eg.db, prim, "SYMBOL", "UNIPROT", multiVals = "first"))
mm <- !is.na(mapIds(org.Mm.eg.db, prim, "SYMBOL", "UNIPROT", multiVals = "first"))

species_summary <- data.table(
  Category   = c("Human-only mapping",
                 "Mouse-only mapping",
                 "Both species mapping",
                 "Neither species mapping"),
  N_proteins = c(sum( hs & !mm),
                 sum(!hs &  mm),
                 sum( hs &  mm),
                 sum(!hs & !mm))
)
species_summary[, Percent := round(100 * N_proteins / sum(N_proteins), 1)]
species_summary
```

The split is approximately **50 % human / 50 % mouse**, not the \~100 % human composition that a human-only FASTA would produce. **Zero proteins map to both species**, ruling out cross-species ambiguity as the explanation — every mouse-mapped accession is unambiguously mouse.

## Evidence 3 — Specific mouse accessions in the output

A handful of representative mouse-only accessions, each annotated by `org.Mm.eg.db` and *not present* in `org.Hs.eg.db`:

```{r, warning=FALSE, message=FALSE}
mouse_examples <- data.table(Protein = prim[!hs & mm])[
  , Mouse_Gene := mapIds(org.Mm.eg.db, Protein, "SYMBOL", "UNIPROT", multiVals = "first")
][
  , Mouse_Description := mapIds(org.Mm.eg.db, Protein, "GENENAME", "UNIPROT", multiVals = "first")
]
head(mouse_examples, 15)
```

Cross-check a single accession on UniProt: e.g. `P12388` resolves to Serpinb2 (Mus musculus); `P48455` to Ppp3cc (Mus musculus); `Q60855` to Ripk1 (Mus musculus). These are unambiguously mouse identifiers in the global UniProt database — they cannot have come from a human-only FASTA.

## Evidence 4 — Parallel `_mouse` report exists

The input directory also contains `20260423_094931_S5-A10_Report_mouse.tsv`, searched against `uniprotkb_proteome_UP000000589_AND_revi_2025_04_18.fasta` (UniProt proteome ID `UP000000589` is *Mus musculus*). The two reports have identical row counts (6 156 104) and were generated from the same raw spectra at adjacent times.

This is consistent with the standard xenograft-proteomics pattern of running two complementary searches — but it does not change the substance of the audit: the report labelled `_human` does not contain only human protein assignments.

```{r, warning=FALSE, message=FALSE}
files <- list.files(input_dir, pattern = "Report_(human|mouse)\\.tsv$",
                    full.names = TRUE)
file.info(files)[, c("size", "mtime")]
```

## Evidence 5 — Mouse report: species composition

If the `_mouse` search used a clean mouse-only FASTA (`UP000000589`), we expect its protein identifications to be overwhelmingly mouse. We apply the same UniProt-mapping check to the unique protein groups in the `_mouse` report.

```{r, warning=FALSE, message=FALSE}
mouse_report <- fread(
  file.path(input_dir, "20260423_094931_S5-A10_Report_mouse.tsv"),
  select = "PG.ProteinGroups"
)

mouse_prim <- str_extract(
  unique(mouse_report$PG.ProteinGroups), "^[^;]+"
)

hs_m <- !is.na(mapIds(org.Hs.eg.db, mouse_prim, "SYMBOL", "UNIPROT", multiVals = "first"))
mm_m <- !is.na(mapIds(org.Mm.eg.db, mouse_prim, "SYMBOL", "UNIPROT", multiVals = "first"))

species_summary_mouse <- data.table(
  Category   = c("Human-only mapping",
                 "Mouse-only mapping",
                 "Both species mapping",
                 "Neither species mapping"),
  N_proteins = c(sum( hs_m & !mm_m),
                 sum(!hs_m &  mm_m),
                 sum( hs_m &  mm_m),
                 sum(!hs_m & !mm_m))
)
species_summary_mouse[, Percent := round(100 * N_proteins / sum(N_proteins), 1)]
species_summary_mouse
```

The `_mouse` report also shows an approximately **50 % human / 50 % mouse** split — mirroring the `_human` report from Evidence 2. This means the mouse FASTA (`UP000000589` by name) is equally contaminated with human sequences. Both searches were therefore run against a concatenated human + mouse database, regardless of the filename used for each.

This result also validates the UniProt accession-mapping approach: the same accessions that appear human in the `_human` report appear human in the `_mouse` report, and vice versa. The 50/50 split is not a mapping artefact but a genuine property of the databases used.

## Conclusion

1.  **Both** search databases — `20K_HUMAN_1GENE_202210.fasta` (labelled human) and `uniprotkb_proteome_UP000000589_...fasta` (labelled mouse) — contain approximately **50 % human and 50 % mouse protein sequences**.
2.  The `_human` and `_mouse` Spectronaut reports are therefore not organism-specific outputs; each is a 50 / 50 chimera of human and mouse identifications.
3.  The most likely explanation is that the STP ran all xenograft samples against a single concatenated human + mouse FASTA and exported two report files with different names, rather than two genuinely separate organism-specific searches.
4.  Downstream analysis must apply explicit species filtering (retain human-mapping accessions only for iPSC-microglia biology, or mouse-mapping accessions only for host-contamination assessment) regardless of which report file is used as input.
