---
title: "4. FASTA Audit — Mislabelled Search Database"
---

## TL;DR

The Spectronaut report delivered as `20260423_094931_S5-A10_Report_human.tsv` was searched against a FASTA file named `20K_HUMAN_1GENE_202210.fasta` — implying a human-only proteome database. **The output contains ~50 % mouse-origin proteins**, which is incompatible with a strictly human FASTA: UniProt accessions are globally unique across organisms, so a clean mouse accession in the output proves the FASTA contains mouse sequences.

The likely facts:
1. The FASTA file is a concatenated **human + mouse** database (appropriate for a xenograft experiment).
2. The filename suggests human-only, which is misleading.
3. Without species filtering, every downstream analysis is contaminated by host (mouse) proteins — this doubles the multiple-testing burden, distorts median normalisation, and confounds the WT vs KO contrast with sample-to-sample host contamination.

This notebook lays out the evidence and recommends post-search species filtering as the fix.

## Setup

```{r setup, message=FALSE}
library(data.table)
library(dplyr)
library(stringr)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(org.Mm.eg.db)

base_dir    <- "/nemo/lab/destrooperb/home/shared/zanettc/dom_proteomics"
input_dir   <- file.path(base_dir, "data", "inputs",
                         "2026_04_23_SORL1_mg_total_proteomics")
results_dir <- file.path(base_dir, "results", "run1")
```

## Evidence 1 — Filename and stated database

The delivered report is named **`20260423_094931_S5-A10_Report_human.tsv`** with sibling files `*.params`, `*.setup.txt`, `*.log.txt`. The setup file declares the search database used.

```{r}
setup_path <- file.path(input_dir, "20260423_094931_S5-A10_Report_human.setup.txt")

fasta_lines <- system2(
  "grep", c("-A1", "Original File", setup_path),
  stdout = TRUE
)
unique(fasta_lines[grepl("Original File", fasta_lines)])
```

The setup file states a single search database: `20K_HUMAN_1GENE_202210.fasta`. By name alone, this implies a Homo sapiens proteome with one canonical isoform per gene (~20 000 entries).

## Evidence 2 — UniProt accession identity is species-specific

UniProt accessions are **globally unique across organisms** — the same accession never refers to two proteins in two species. Therefore, if the output of a search against a strictly human FASTA contains an accession that, in the UniProt knowledge base, is annotated to a non-human organism, the FASTA cannot have been strictly human.

We test this against the full set of unique protein groups in the MSstats output by mapping the primary accession of each group to (a) `org.Hs.eg.db` and (b) `org.Mm.eg.db`.

```{r}
res <- fread(file.path(results_dir, "Full_GroupComparison_Results.csv"))
prim <- str_extract(res$Protein, "^[^;]+")  # first accession in semicolon-separated group

hs <- !is.na(mapIds(org.Hs.eg.db, prim, "SYMBOL", "UNIPROT", multiVals = "first"))
mm <- !is.na(mapIds(org.Mm.eg.db, prim, "SYMBOL", "UNIPROT", multiVals = "first"))

species_summary <- data.table(
  Category   = c("Human-only mapping",
                 "Mouse-only mapping",
                 "Both species mapping",
                 "Neither species mapping"),
  N_proteins = c(sum( hs & !mm),
                 sum(!hs &  mm),
                 sum( hs &  mm),
                 sum(!hs & !mm))
)
species_summary[, Percent := round(100 * N_proteins / sum(N_proteins), 1)]
species_summary
```

The split is approximately **50 % human / 50 % mouse**, not the ~100 % human composition that a human-only FASTA would produce. **Zero proteins map to both species**, ruling out cross-species ambiguity as the explanation — every mouse-mapped accession is unambiguously mouse.

## Evidence 3 — Specific mouse accessions in the output

A handful of representative mouse-only accessions, each annotated by `org.Mm.eg.db` and *not present* in `org.Hs.eg.db`:

```{r}
mouse_examples <- data.table(Protein = prim[!hs & mm])[
  , Mouse_Gene := mapIds(org.Mm.eg.db, Protein, "SYMBOL", "UNIPROT", multiVals = "first")
][
  , Mouse_Description := mapIds(org.Mm.eg.db, Protein, "GENENAME", "UNIPROT", multiVals = "first")
]
head(mouse_examples, 15)
```

Cross-check a single accession on UniProt: e.g. `P12388` resolves to Serpinb2 (Mus musculus); `P48455` to Ppp3cc (Mus musculus); `Q60855` to Ripk1 (Mus musculus). These are unambiguously mouse identifiers in the global UniProt database — they cannot have come from a human-only FASTA.

## Evidence 4 — Parallel `_mouse` report exists

The input directory also contains `20260423_094931_S5-A10_Report_mouse.tsv`, searched against `uniprotkb_proteome_UP000000589_AND_revi_2025_04_18.fasta` (UniProt proteome ID `UP000000589` is *Mus musculus*). The two reports have identical row counts (6 156 104) and were generated from the same raw spectra at adjacent times.

This is consistent with the standard xenograft-proteomics pattern of running two complementary searches — but it does not change the substance of the audit: the report labelled `_human` does not contain only human protein assignments.

```{r}
files <- list.files(input_dir, pattern = "Report_(human|mouse)\\.tsv$",
                    full.names = TRUE)
file.info(files)[, c("size", "mtime")]
```

## Conclusion

1. The file `20K_HUMAN_1GENE_202210.fasta` used for the `_human` Spectronaut search **contains both human and mouse protein sequences**, despite its name.
2. The proteomics facility's deliverable is misleading: a user taking the filename at face value would assume a human-only analysis, but every analysis run on this output is a 50 / 50 chimera unless mouse proteins are explicitly filtered post-search.
3. For a xenograft experiment a mixed FASTA is the correct setup. The issue is the **labelling and documentation**, not the search itself.

## Recommendations

- **Immediate (this project):** filter mouse-origin proteins from the MSstats input prior to `dataProcess` so normalisation, statistical testing, and FDR are computed on human proteins only. This is implemented in notebook `02_msstats.qmd`.
- **For future deliverables:** request the facility to either rename the FASTA file to reflect its true content (e.g. `HUMAN_MOUSE_combined_*.fasta`) or supply an unambiguously human-only FASTA for searches whose stated purpose is human-only.
- **Reproducibility:** retain both `_human` and `_mouse` reports for transparency. The `_mouse` report is searched against a verifiable mouse proteome (`UP000000589`) and could support a future host-response analysis if needed.
